---
title: "Drugs and Drug Classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drugs and Drug Classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.height = 13,fig.width = 13)
```

# Introduction

```{r}
library(PDSdatabase)
library(rlang)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggthemes)

theme_set(
    ggthemes::theme_tufte() + theme(text = element_text(size=16,face="bold"))
)
```

This vignette showcases the reported drugs and drug classes in the PDSdatabase:

- How many drug compounds are listed in reports

- How many classes of drugs are listed in reports

- How many drugs and classes of drugs are listed in reports

- What are the most frequent drugs and classes of drugs listed in reports across child development stages


```{r}

download_sqlite_db() #downloads to current working directory

```


# Connect to the database:

```{r}

con <- connect_sqlite_db()

```


# How many drug compounds are listed in reports?

The field `atc_concept_id` is the concept adentifier for each drug's ingredient encoded in the ATC vocabulary. Below is the number of distinct drug ingredients or compounds:

```{r}

tbl(con,"ade_raw") %>% 
    distinct(atc_concept_id) %>% 
    count()

```

The number of drug compounds across reports:

```{r,fig.height=5,fig.width=8}

tbl(con,"ade_raw") %>% 
    distinct(safetyreportid,atc_concept_id) %>% 
    group_by(safetyreportid) %>% 
    count(name = "Ndrugs") %>% 
    group_by(Ndrugs) %>% 
    count(name="Nreports") %>% 
    collect() %>% 
    ggplot(aes(Ndrugs,Nreports)) +
    geom_bar(stat="identity",color="black",fill="gray") +
    labs(x="Number of drugs listed in reports",
         y="Number of reports") +
    scale_y_continuous(trans="log10",labels=scales::comma) 

```

# How many classes of drugs are listed in reports?

```{r,fig.height=8}
tbl(con,"ade_raw") %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        tbl(con,"drug") %>% 
            distinct(atc_concept_id,atc1_concept_name) %>% 
            collect(),
        by="atc_concept_id"
    ) %>% 
    group_by(atc1_concept_name) %>% 
    summarise(
        N = n_distinct(safetyreportid)
    ) %>% 
    ggplot(aes(N,atc1_concept_name)) +
    geom_bar(stat="identity",color="black",fill="gray") +
    labs(x="Number of reports",y="")
```

# What are the most frequent drugs and classes of drugs listed in reports across child development stages?

```{r}

scale_y_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
}

reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new <- paste(x, within, sep = sep)
  stats::reorder(new, by, FUN = fun)
}

drug_tbl <- 
    tbl(con,"drug") %>% 
    distinct(atc_concept_id,atc_concept_name,
             !!!rlang::syms(paste0("atc",1:4,"_concept_name"))
             ) %>% 
    collect() %>% 
    rename(atc5_concept_name = atc_concept_name) %>% 
    pivot_longer(
        cols=paste0("atc",1:5,"_concept_name"),
        names_to = "atc_class",
        values_to = "value"
    ) 

drug_tbl

```

## In all childhood

```{r}

tmp <- tbl(con,"ade_raw") %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports from birth through 21 years old") +
    ylab("")

```

## Birth through 1 month olds - Term Neonates

```{r}

stage="term_neonatal"

tmp <- 
    tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in term neonates") +
    ylab("")

```

## 1 month through 1 year olds - Infants

```{r}

stage="infancy"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in infants") +
    ylab("")

```

## 1 year through 2 years old - Toddlers

```{r}

stage="toddler"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in toddlers") +
    ylab("")

```

## 2 years through 5 years old - Early childhood

```{r}

stage="early_childhood"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in early childhood") +
    ylab("")

```

## 5 years through 11 years old - Middle childhood

```{r}

stage="middle_childhood"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in middle childhood") +
    ylab("")

```


## 11 years through 17 years old - Early adolescence

```{r}

stage="early_adolescence"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in early adolescence") +
    ylab("")

```

## 17 years through 21 years old - Late adolescence

```{r}

stage="late_adolescence"

tmp <- tbl(con,"ade_raw") %>% 
    filter(nichd==!!stage) %>% 
    distinct(safetyreportid,
           atc_concept_id) %>% 
    collect() %>% 
    inner_join(
        drug_tbl,
        by="atc_concept_id"
    ) %>% 
    group_by(value) %>% 
    summarise(
        Nreports = n_distinct(safetyreportid)
        ) %>% 
    inner_join(
        drug_tbl %>% 
            distinct(atc_class,value),
        by="value"
    ) %>% 
    arrange(desc(Nreports)) %>% 
    group_by(atc_class) %>% 
    slice_head(n=10)

tmp %>% 
    ggplot(aes(Nreports,reorder_within(value,Nreports,atc_class))) +
    geom_bar(stat="identity") +
    scale_y_reordered() +
    facet_wrap(~atc_class,ncol=1,scales="free_y") +
    xlab("Number of safety reports in late adolescence") +
    ylab("")

```


# Disconnect database:

```{r}

disconnect_sqlite_db(con)

```

